---
title: "Sandy_MA_2022"
author: "Sandy Hickson"
date: '2022-08-28'
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

#Sex-bias Classifications
Here, we classify sex-bias (M,F,U) for each coexpressed gene within the mutation accumulation (MA) and ancestral treatment (ANC) treatments, using the log2 fold-change difference in expression (exp) between males and females.
this code chunck produces 'SexBias_TMM2.txt', raw data for this code chunk is published at 'X'.
```{r, echo=FALSE}
library(data.table)
library(tidyverse)
library(doParallel)

#Read in raw Data (published at: 'URL')
dat <- fread("../data/normData_tall_coexpressed2_incAncestor_TMM_strictCPM2.txt") %>% 
  filter(treatment == "Ancestor" |
          treatment == "MA" ) %>% 
  filter(!is.na(CHROM)) %>% 
  mutate(Recip = case_when(str_detect(sampleID, "FM") ~ "FM",
                        str_detect(sampleID, "MM") ~ "MM",
                        str_detect(sampleID, "MF") ~ "MF",
                        str_detect(sampleID, "FF") ~ "FF")) %>% 
  filter(!(CHROM == "X" &  Recip == "MM"))


#DoParallel
detectCores()
registerDoParallel(6)
start <- proc.time()[3]
# 
SexBias <- foreach(i = 1:n_distinct(dat$geneID), .combine = rbind,
                    .packages = 'tidyverse',
                    .errorhandling = "remove") %dopar% {

genename = unique(dat$geneID)[i]
  
  #Sex-bias in Anc  
  dat %>% 
    filter(geneID == genename) -> test
    
     test[grepl("226", test$line),] -> test
     
     model <- lm(as.numeric(expr) ~ offspringSex + Recip, data=test)
        
        anova(model) -> pvalues
        
        pvalues$"Pr(>F)"[1] -> pSex
        
        as.data.frame(cbind(genename, pSex)) -> pdata
        
        test %>% filter(offspringSex == "female") -> tempF
        
        test %>% filter(offspringSex == "male") -> tempM
        
        mean(tempF$expr) -> meanF
        
        mean(tempM$expr) -> meanM
        
        meanM - meanF -> meandif
        
        ifelse(meandif>1, "M", ifelse(meandif>-1 & meandif<1, "U", "F")) -> bias
        
        
        as.data.frame(cbind(meanM, meanF, meandif, bias)) -> foldchange

        
 ##################################################
        #Sex-bias in MA

        filter(dat, geneID == genename) -> test_MA
        
        test_MA[!grepl("226", test_MA$line),] -> test_MA
        
        model_MA <- lm(as.numeric(expr) ~ offspringSex + Recip, data=test_MA)
        
        anova(model_MA) -> pvalue_MA
        
        pvalue_MA$"Pr(>F)"[1] -> pSex_MA
        
        test_MA %>% filter(offspringSex == "female") -> tempF_MA
        
        test_MA %>% filter(offspringSex == "male") -> tempM_MA
        
        mean(tempF_MA$expr) -> meanF_MA
        
        mean(tempM_MA$expr) -> meanM_MA
        
        meanM_MA - meanF_MA -> meandif_MA
        
        ifelse(meandif_MA>1, "M", ifelse(meandif_MA>-1 & meandif_MA<1, "U", "F")) -> bias_MA
        
        as.data.frame(cbind(pSex_MA, meanM_MA, meanF_MA, meandif_MA, bias_MA)) -> foldchange_MA
        
        
        
        temp <- cbind(pdata, foldchange, foldchange_MA)
        temp
        
 }

time <- proc.time()[3] - start

SexBias %>% 
  mutate_at(c(2:5,7:10), as.numeric) -> SexBias

SexBias$pSex_BH <- p.adjust(SexBias$pSex, method = "BH")
SexBias$pSex_BH_MA <- p.adjust(SexBias$pSex_MA, method = "BH")

SexBias$meandifdif <- abs(SexBias$meandif_MA) - abs(SexBias$meandif)

SexBias$meandifdiv <- (SexBias$meandif_MA)/(SexBias$meandif)

SexBias$meandifdiffoSexBiasmales <- abs(SexBias$meanM_MA) - abs(SexBias$meanM)

SexBias$meandifdiffoSexBiasfemales <- abs(SexBias$meanF_MA) - abs(SexBias$meanF)

SexBias$meandifdifmoSexBiase <- (SexBias$meandifdiffoSexBiasmales) - (SexBias$meandifdiffoSexBiasfemales)

##deltaM

SexBias %>% 
  mutate(deltaM_M=(((meanM_MA - meanM)/28)/meanM)*100) %>% 
  mutate(deltaM_F=(((meanF_MA - meanF)/28)/meanF)*100) ->
  SexBias

##Filter out false postives for bias
SexBias %>% 
  mutate(bias = case_when(pSex_BH > 0.05 ~ "U",
                          TRUE ~ bias),
         bias_MA = case_when(pSex_BH_MA > 0.05 ~ "U",
                          TRUE ~ bias_MA)) ->
  SexBias

write.table(SexBias, "../data/SexBias_TMM2.txt")

```

#Statistical Analysis
##Sex-bias: Count Data
Here, we use Chi-squared tests to detect changes in the number of genes 
within each sex biased class.
```{r}
library(data.table)
library(tidyverse)

CHROM <- fread("../data/scaff2gene.txt") %>% 
  select(!V1) %>% 
  mutate(CHROM_SH=case_when(CHROM=="X" ~ "X",
                            CHROM=="NA" ~ "NA",
                            TRUE ~ "Auto")) %>% 
  select(geneID,CHROM, CHROM_SH)

SexBias <- fread("../data/SexBias_TMM2.txt") %>% 
  select(!V1)

############################################
### Sex Bias Count Data
#
SexBias %>% 
  count(bias_MA) %>% 
  rename(MA=n, bias = bias_MA) %>%   inner_join(rename(count(SexBias, bias), ANC = n), by = "bias") -> bias_count


SexBias %>% 
  mutate(bias_bin = case_when(bias == "U" ~ "U",
                              TRUE ~ "B"),
         bias_bin_MA = case_when(bias_MA == "U" ~ "U",
                              TRUE ~ "B")) -> biasBin

  

#########################
#Comparing all classes
chisq <- chisq.test(data_frame(ANC = bias_count$ANC, MA = bias_count$MA), correct = FALSE)
chisq
chisq$p.value

#########################
#Comparing bias vs unbiased
SexBias %>% 
  mutate(bias_bin = case_when(bias == "U" ~ "U",
                              TRUE ~ "B"),
         bias_bin_MA = case_when(bias_MA == "U" ~ "U",
                              TRUE ~ "B")) -> biasBin

All_ANC <- count(biasBin, bias_bin) 
All_MA <- count(biasBin, bias_bin_MA)

cbind(All_ANC$n, All_MA$n) -> SB_All 
colnames(SB_All)<-  c("ANC", "MA")
rownames(SB_All)<-  c("B", "U")

chisq2 <- chisq.test(SB_All, correct = FALSE)
chisq2
chisq2$p.value
chisq2$observed
chisq2$expected

#########################
#within sex-biased genes
#Male-bias
MB_ANC <- filter(SexBias, bias == "M") %>% 
  count(bias) 

MB_MA <- filter(SexBias, bias_MA == "M") %>% 
  count(bias_MA)

cbind(MB_ANC$n, MB_MA$n) %>% 
  set_names("ANC", "MA") %>% 
  data_frame() -> MB

chisq3 <- chisq.test(MB, correct = FALSE)
chisq3
chisq3$p.value
chisq3$observed
chisq3$expected


#Female-bias
FB_ANC <- filter(SexBias, bias == "F") %>% 
  count(bias) 

FB_MA <- filter(SexBias, bias_MA == "F") %>% 
  count(bias_MA)

cbind(FB_ANC$n, FB_MA$n) %>% 
  set_names("ANC", "MA") %>% 
  data_frame() -> FB

chisq3 <- chisq.test(FB, correct = FALSE)
chisq3
chisq3$p.value
chisq3$observed
chisq3$expected

#########################
#between unbiased genes
SB_ANC <- filter(SexBias, bias != "U") %>% 
  count(bias) 

SB_MA <- filter(SexBias, bias_MA != "U") %>% 
  count(bias_MA)

cbind(SB_ANC$n, SB_MA$n) -> SB 
colnames(SB)<-  c("ANC", "MA")
rownames(SB)<-  c("F", "M")

chisq3 <- chisq.test(SB, correct = FALSE)
chisq3
chisq3$p.value
chisq3$observed
chisq3$expected

```

##Sex-bias: Effect Size data
Here, we use a Permutation-based ANOVA and Wilcoxon rank sum tests, which are both a non-parametric tests.
```{r}
library(data.table)
library(tidyverse)
library(permuco)

setwd("C:/Users/Sandy/OneDrive - The University of Queensland/MutationBias_2022/src")

SexBias <- fread("../data/SexBias_TMM2.txt") %>% 
  select(!V1)

SexBias %>% 
  mutate(bias_bin = case_when(bias == "U" ~ "U",
                              TRUE ~ "B"),
         bias_bin_MA = case_when(bias_MA == "U" ~ "U",
                              TRUE ~ "B")) -> biasBin

#######################################################
#Permutation based anova
SexBias %>% 
  select(geneID, bias, sexbias, sexbias_MA) %>% 
  pivot_longer(cols = c("sexbias", "sexbias_MA"), 
               names_to = "treat", values_to = "sexbias") -> bias_perm

aovperm(abs(sexbias) ~ treat*bias, np= 10000, data = biasBin_perm)
  
bias_perm %>%
  group_by(bias, treat) %>% 
  summarise(median(abs(sexbias))) %>% 
  pivot_wider(names_from = "treat", values_from = "median(abs(sexbias))") %>% 
  mutate(dif = sexbias - sexbias_MA)#?!?

###Post-hoc testing
##Wilcoxon rank sum tests
#All Genes
test1 <- wilcox.test(abs(SexBias$sexbias), abs(SexBias$sexbias_MA), 
                     alternative = "two.sided")
test1
test1$p.value

#SB genes
SB_test <- wilcox.test(abs(filter(biasBin, bias_bin == "B")$sexbias), 
                     abs(filter(biasBin, bias_bin == "B")$sexbias_MA), 
                     alternative = "two.sided")
SB_test
SB_test$p.value
#UB genes
test2 <- wilcox.test(abs(filter(SexBias, bias == "U")$sexbias), 
                     abs(filter(SexBias, bias == "U")$sexbias_MA), 
                     alternative = "two.sided")
test2
test2$p.value

#MB 
test3 <- wilcox.test((filter(SexBias, bias == "M")$sexbias), 
                     (filter(SexBias, bias == "M")$sexbias_MA), 
                     alternative = "two.sided")
test3
test3$p.value

#FB
test4 <- wilcox.test((filter(SexBias, bias == "F")$sexbias), 
                     (filter(SexBias, bias == "F")$sexbias_MA), 
                     alternative = "two.sided",)
test4
test4$p.value
```

##Mutational Bias: Significance testing
```{r}
library(data.table)
library(tidyverse)
library(permuco)


SexBias <- fread("../data/SexBias_TMM2.txt") %>% 
   select(!V1)
  

SexBias %>%
  select(geneID, bias, deltaM_M, deltaM_F, sexbias, sexbias_MA, pSex_BH, CHROM, CHROM_SH) %>% 
  pivot_longer(cols = c("deltaM_M", "deltaM_F"), 
               names_to = c(".value", "Sex"), names_sep = "_") ->
  deltaM

#######################################################
#Permutation based anova
Aov_dM <-aovperm(deltaM ~ Sex*bias, np= 10000, 
                 data = filter(deltaM, CHROM_SH == "Auto"))
Aov_dM#?!?

###Post-hoc testing
##Mann-Whitney Test: Autosomal genes
#All Genes
Test1 <- wilcox.test((SexBias$deltaM_M), (SexBias$deltaM_F), 
                     alternative = "two.sided")
Test1
Test1$p.value

#Male-bias
Test2 <- wilcox.test(filter(SexBias, bias == "M", CHROM_SH == "Auto")$deltaM_M,
                     filter(SexBias, bias == "M", CHROM_SH == "Auto")$deltaM_F, 
                     alternative = "two.sided")
Test2
Test2$p.value

#Female-bias
Test3 <- wilcox.test(filter(SexBias, bias == "F", CHROM_SH == "Auto")$deltaM_M,
                     filter(SexBias, bias == "F", CHROM_SH == "Auto")$deltaM_F, 
                     alternative = "two.sided")
Test3
Test3$p.value

#Unbiased
Test4 <- wilcox.test(filter(SexBias, bias == "U", CHROM_SH == "Auto")$deltaM_M,
                     filter(SexBias, bias == "U", CHROM_SH == "Auto")$deltaM_F, 
                     alternative = "two.sided")
Test4
Test4$p.value


##Mann-Whitney Test
#All Genes
Test1 <- wilcox.test((SexBias$deltaM_M), (SexBias$deltaM_F), 
                     alternative = "two.sided")
Test1
Test1$p.value

#Male-bias
Test2 <- wilcox.test(filter(SexBias, bias == "M", CHROM_SH == "X")$deltaM_M,
                     filter(SexBias, bias == "M", CHROM_SH == "X")$deltaM_F, 
                     alternative = "two.sided")
Test2
Test2$p.value

#Female-bias
Test3 <- wilcox.test(filter(SexBias, bias == "F", CHROM_SH == "X")$deltaM_M,
                     filter(SexBias, bias == "F", CHROM_SH == "X")$deltaM_F, 
                     alternative = "two.sided")
Test3
Test3$p.value

#Unbiased
Test4 <- wilcox.test(filter(SexBias, bias == "U", CHROM_SH == "X")$deltaM_M,
                     filter(SexBias, bias == "U", CHROM_SH == "X")$deltaM_F, 
                     alternative = "two.sided")
Test4
Test4$p.value

```


